---
output: github_document
bibliography: "/home/kd/PARA/Notes/My Library.bib"
refrence-section-title: Refrences
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# liaendmembers

<!-- badges: start -->
[![DOI](https://zenodo.org/badge/DOI/10.5281/zenodo.16909256.svg)](https://doi.org/10.5281/zenodo.16909256)
<!-- badges: end -->

This package provides convenient tools for provenance analysis using lead isotope data. Specifically, the software enables researchers to:

- Identify endmembers for provenance analysis using Principal Component Analysis [@A.D.G+24] and group components by their distribution along the geochron slope [@E.E.Y+19].

- Calculate nearest distances between sample and reference data within ore isotope databases using Euclidean metrics with or without mass-dependent fractionation corrections [@A.D.B+24].

- Determine probabilities of samples originating from specific subregional groups as defined by established isotope databases [@S.K.D+26].

## Installation

You can install the development version of liaendmembers like so:

```
devtools::install_github("karan3242/liaendmembers")
```

## Datasets

This package comes with Lead Isotope analyse of Iron Age silver hoards from Tel
Dor, Akko, Arad, Ein Hofez and Eshtemoa in Israel [@E.E.Y+19],

```{r data}
head(liaendmembers::tel_dor)
```

along with a pretrained preovence `xgb.Booster` model for prevenance prediction.

```{r data}
head(liaendmembers::model, 3)
```

## Example

This is a basic example which shows you how to solve a common problem:

```{r example}
library(liaendmembers)
## Using the endmemebrs function
data("tel_dor")
end_members <- endmembers(
     tel_dor,
     colnames(tel_dor),
     tolerance = c(0.01, 0.01),
     clamp = c(Inf, Inf)
)
## Print summary of the liaendmembers object
summary(end_members)

## Use predictive model to find the probablity of provenance

isoprov_predict(end_members, model)
```
